Notes and other text

### 2023-04-22 doc/manuscript/translocation.qmd
Add final paragraphs to Discussion, edits elsewhere

Emerging infectious diseases are increasing [@daszak2000] and pose a severe threat to human, livestock, and wildlife populations.
In wildlife populations, disease is responsible for dramatic declines, extirpations, and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Although the dramatic and immediate nature of disease-induced declines in wildlife has led to intensive focus on documenting and understanding disease impacts [@langwig2015], this information is generally insufficient to allow the development of effective methods to recover disease-impacted species [@scheele2021; @fellers2007].
Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
Unfortunately, given the nascent nature of wildlife recoveries in the presence of disease, we know far less about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
At a time when novel infectious diseases are emerging at an unprecedented rate [@fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022].
However, linking these selective responses directly to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
In addition to promoting recovery of remnant populations after disease-induced declines, the resistant/tolerant individuals <!--# How best to refer to resistance/tolerance? --> in recovering populations can also be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survive following disease emergence evolve increased resistance/tolerance.
Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and are maintained across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the Sierra Nevada mountains [California, USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolutionary rescue in response to disease allows the successful reestablishment of extirpated MYL frog populations, and that these populations have a high probability of long-term persistence despite ongoing Bd infection.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in recovering populations [@knapp2016].
We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize population models that extend our inferences of population viability well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolutionary rescue and the reintroduction of resistant animals can allow the large-scale recovery of disease-impacted species, with potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-selectionresults A: SI) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@knapp2011, @vredenburg2010].

We used two complementary approaches to identify regions of the genome under selection.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
Second, we used a splined window approach to identify larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).

The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults C, D: SI, SI Dataset 1).
One of the 7 identified genes did not have an associated annotation.
Frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.

The splined window analysis identified 58 outlier regions for *F~ST~* and 33 outlier regions for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2).
The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and four occur in the extended MHC Class III region (HSP70, LSM2, VARS <!--# Check on number of genes, add missing gene here -->) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes.

In summary, our genomics results show that recovering populations have signatures of strong selection on multiple regions of the genome.
Although the functions of all associated genes were not clearly interpretable, regions of the genome under selection in recovering populations included those associated with resistance/tolerance in other amphibian systems, such as the MHC complex.

<!--# Based on Fig. 2, there are numerous outlier windows. How were the specific windows mentioned in above and in Fig. 2 C, D chosen? Seems important to mention to avoid the perception that mentioned patterns were chosen haphazardly or cherry-picked based on unknown criteria. -->

@@

In this section, we describe the outcome of *R. sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) [@knapp2016] and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@joseph2018; @seddon2014].
The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).

Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].

@@

Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to recruitment probability $\omega$ and yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$ showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).

We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; \@fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 - 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).

In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
In addition, five of the seven populations translocated after 2013 were predicted to have long-term viability, compared to only one of five populations translocated prior to 2013 (@tbl-survival-earlylate).
Thus, there is some trend in our improved ability to select appropriate habitat to promote long-term viability of translocated, resistant populations.

@@

Disease-induced population declines are impacting global biodiversity [@daszak2000], and strategies to recover affected species are lacking.
Recovery in this context is likely to be particularly challenging because typically recovery will need to occur in the presence of ongoing disease.
One potential strategy, and the one we employ in the current study, utilizes populations of resistant/tolerant individuals as a source of animals with which to reestablish extirpated populations in the presence of disease.
Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to recover populations.
Here, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014], obtaining strong evidence in wild animal populations for each of the associated causal links (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
Nonetheless, in the following summary of our results, we use evolutionary rescue as an informative, but tentative framework.

Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter - before translocation].
These differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.

>>>>

Emerging infectious diseases pose a severe threat to wildlife populations [@daszak2000], and have caused dramatic declines and even extinctions in mammals, birds, amphibians, echinoderms, and other taxa [@hewson2014; @samuel2015; @scheele2019; @cunningham2021].
Importantly, some populations previously decimated by disease have subsequently recovered to varying degrees [@newell2013; @voyles2018; @knapp2016] and might provide critical insights for species conservation efforts.
Unfortunately, given the rarity of wildlife recoveries in the presence of disease, we know little about the mechanisms driving population recovery [@brannelly2021; @russell2020] and how these mechanisms can be leveraged to re-establish host populations across large spatial scales [@mendelson2019].
At a time when novel infectious diseases are emerging at an unprecedented rate [@daszak2000; @fisher2012], this knowledge gap has far-reaching consequences for global biodiversity.

Host evolution following disease-induced declines is one mechanism by which hosts can recover [@carlson2014; @searle2020].
"Evolutionary rescue" describes rapid evolutionary change that increases the frequency of adaptive alleles and thereby restores positive population growth and prevents extinction [@carlson2014].
Evolutionary changes following disease-induced declines are known from a diversity of wildlife populations [@savage2016; @epstein2016; @gignoux-wolfsohn2021; @holland2022], and often result in increased host resistance and/or tolerance.
However, linking these selective responses to population recoveries - as necessary for evolutionary rescue - is difficult and remains rare.
In addition to promoting recovery of disease-impacted populations, the resistant/tolerant individuals in recovering populations can, in theory, be used to reestablish populations that were extirpated following disease emergence [e.g., @mendelson2019].
To date, the use of resistant/tolerant individuals for species recovery is rare [but see @joseph2018], and studies that evaluate its potential utility are critically needed.

There are four conditions that will often determine the feasibility of using individuals from populations that are recovering following disease-induced declines to reestablish extirpated populations.
First, populations that survive following disease emergence evolve increased resistance/tolerance.
Second, increased resistance/tolerance contributes to higher individual fitness in the presence of disease and allows positive population growth.
Third, the fitness benefits gained through selection for increased resistance are maintained beyond the natal environment and across generations, allowing for the establishment of reintroduced populations.
And fourth, reintroduced populations show long-term growth and persistence.
To date, few of these conditions are documented for any single species impacted by disease, and as a result the reintroduction of resistant hosts as a means to recover species of conservation concern remains largely an untested idea [@scheele2021; @mendelson2019].

The mountain yellow-legged (MYL) frog, comprised of the sister species *Rana muscosa* and *Rana sierrae* [@vredenburg2007], is emblematic of the unprecedented global declines of amphibians caused by the recently-emerged amphibian chytrid fungus [*Batrachochytrium dendrobatidis,* @scheele2019].
Once the most common amphibian in the high elevation portion of the California's Sierra Nevada mountains [USA, @grinnell1924], this frog has disappeared from more than 90% of its historical range during the past century [@vredenburg2007].
Due to the severity of its decline and the increasing probability of extinction, both species are now listed as "endangered" under the U.S.
Endangered Species Act.
In the Sierra Nevada, this decline was initiated by the introduction of non-native trout into the extensive historically-fishless region [@bradford1989; @knapp2000], and the arrival of *B. dendrobatidis* (Bd) in the 1960s and its subsequent spread [@vredenburg2019] caused additional widespread population extirpations [@vredenburg2010; @rachowicz2006].
These Bd-caused declines are fundamentally different from the fish-caused declines because removal of fish populations is feasible and results in the rapid recovery of frogs [@knapp2007; @vredenburg2004].
In contrast, Bd appears to persist in habitats even in the absence of amphibian hosts [@walker2007], and therefore represents a long-term alteration of invaded ecosystems that amphibians will need to overcome to reestablish populations.

Despite the catastrophic impact of Bd on MYL frogs in which most Bd-naive populations are driven to extirpation following Bd arrival [@vredenburg2010], some populations have persisted after epizootics [@briggs2010] and are now recovering [@knapp2016].
Frogs in these recovering populations show reduced susceptibility to Bd infection, consistent with selection for more resistant/tolerant genotypes [@knapp2016].
Furthermore, exploratory efforts to reintroduce frogs collected from recovering populations to sites from which frogs were previously extirpated indicate that population establishment is possible despite ongoing Bd infection [@joseph2018].

In the current study, we combine results from genomic analyses, a 15-year landscape-scale frog reintroduction effort, and detailed population models to provide an unprecedented example of how evolution in response to disease allows the successful reestablishment of extirpated MYL frog populations.
Specifically, we used exome capture methods to show that MYL frogs have undergone substantial evolutionary change associated with Bd exposure and that may provide a genetic basis for the observed increased resistance/tolerance to Bd infection observed in naturally recovering populations [@knapp2016].
We then describe the reintroduction of these resistant frogs to 24 sites and the dynamics and drivers of population establishment using results from long-term capture-mark-recapture (CMR) surveys.
These data are also used to parameterize a population model that extends our inferences of population establishment well beyond the multi-year extent of the CMR data.
Collectively, these results provide one of the first examples of how evolution and the reintroduction of resistant/tolerant animals can allow the large-scale recovery of disease-impacted species.
This example exhibits several of the key characteristics of evolutionary rescue, and has potentially broad implications for the many taxa worldwide that are threatened with extinction by emerging diseases.

@@

To determine whether frog populations show an evolutionary response to Bd, we compared genomes of frogs in populations with different histories of Bd exposure.
Specifically, we compared frog genomes sampled in four populations that have not yet experienced a Bd-caused epizootic ("naive") [@zhou2015] versus in five populations that experienced a Bd epizootic during the past several decades and have since recovered to varying degrees ("recovering"; @fig-allelefrequencies) [@knapp2016; @vredenburg2010].
These recovering populations persist in an enzootic state [@briggs2010], characterized by high Bd prevalence and, in adults, moderate Bd infection intensities ("load") that are typically well below the level expected to cause mortality [@vredenburg2010].

We conducted a principal component analysis (PCA) of the genomic data to describe the association between sampled populations, and then used two complementary approaches to identify regions of the genome under selection.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and specific single nucleotide polymorphisms (SNPs).
Second, we used a splined window analysis to identify larger genomic regions showing differences between population types in *F~ST~* and nucleotide diversity ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).

Individual frogs clustered into 3 separate groups in principal component space (@fig-selectionresults A: SI), and clusters reflected the strong signature of isolation by distance that is characteristic of MYL frogs [@rothstein2020; @poorten2017].
Importantly, each cluster contained at least one population from both the naive and recovering groups, allowing us to distinguish between (i) allelic associations of individuals sampled in naive versus recovering populations, and (ii) allelic associations resulting from population structure and genetic drift.

Results from the SNP and splined window analyses show that recovering populations have signatures of strong selection on multiple regions of the genome.
The SNP analysis identified 11 outlier SNPs from 7 distinct genes across 4 contigs (@fig-selectionresults B, C: SI, SI Dataset 1).
One of the 7 identified genes did not have an associated annotation.
Frequency differences between the naive and recovering populations ranged from 41% to 86%.
Most of these SNPs showed frequency differences in only a subset of the sampled populations (@fig-allelefrequencies A, B), but the outlier SNP in the RIN3 gene showed consistent differences in frequencies across all populations (@fig-allelefrequencies C).
This is suggestive of strong selection at this locus associated with exposure of frogs to Bd.

The splined window analysis identified 58 outlier regions for *F~ST~* and 33 outlier regions for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and 2 of these shared regions also contained outlier SNPs.
A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2).
The largest $\pi_{diff}$, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains approximately 500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had lower $\pi_{diff}$ in the recovering populations, consistent with balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier window to the VARS SNP) contained a total of 8 annotated genes.
In *Xenopus*, five of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1, CCHCR1, TCF19) and three occur in the extended MHC Class III region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, this region under selection is part of the extended MHC Class I and III complex and shows synteny with other amphibian genomes (@fig-synteny-plot SI).

@@

In this section, we describe the outcome of *R. sierrae* reintroductions conducted in Yosemite National Park as part of a 15-year effort to reestablish populations of this endangered species (`r min(d1$year)`--`r max(d1$year)`).
Each of the `r n_distinct(d1$site_id, d1$date)` reintroductions involved collection of adult frogs from 1 of `r n_distinct(d1$donor)` recovering, Bd-positive "donor" populations (@fig-yosemap SI) and translocating them to 1 of `r n_distinct(d1$site_id)` nearby recipient sites [@seddon2014].
The donor populations included 2 of the 5 recovering populations in the frog evolution study described above (@fig-allelefrequencies A: population 1 and 4).
Following translocation, adult survival and recruitment of new adults were estimated from CMR surveys and counts of tadpoles and juveniles were obtained from visual encounter surveys (VES).

Estimates of frog survival during the 1 year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens SI).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (@tbl-survival-earlylate SI), which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter SI), and loads were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale) [@vredenburg2010; @joseph2018].

@@

Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values SI).
Given observed yearly adult survival probabilities (from analysis of CMR data) and a baseline yearly recruitment probability $\omega$ of 0.3 (i.e., the probability that, following survival, juveniles successfully recruit to the adult class), our stage-structured model predicted that six of twelve translocated populations should experience a long-run growth rate $\lambda$ greater than 1 in the presence of Bd (@fig-viability A; median predicted $\lambda$ ranges from 1.15-1.36 for these six populations).
These six populations all had observed yearly adult survival greater than 0.5.
When we reduced the probability of successful recruitment $\omega$ to 0.1 (a highly conservative estimate), long-run growth rate of five of these six populations was still greater than 1 (@fig-viability A).
In contrast, for the other six translocated populations, where yearly adult survival was substantially less than 0.5, our model predicted that it was unlikely that long-run growth rate in these populations was greater than 1, even when the probability of successful recruitment to the adult life stage was near 0.4 (@fig-viability A).
A sensitivity analysis of long-run growth rate $\lambda$ to yearly adult survival probability of naturally recruited individuals $\sigma_{A_R}$, recruitment probability $\omega$, and female fecundity *F* showed that $\lambda$ was most sensitive to small changes in $\sigma_{A_R}$ (@fig-viability-supp A).
Note that we assumed $\sigma_{A_R}$ was equivalent to the survival probability of translocated adults $\sigma_{A_T}$, except for three populations where we had \> 5 years of CMR data and recovering populations that together allowed us to estimate $\sigma_{A_R}$ and $\sigma_{A_T}$ separately (@fig-compare_surv_probs SI).

We then incorporated demographic stochasticity (i.e., probabilistic births and deaths of individuals in a population) and environmental variability in the probability of juvenile recruitment to the adult class $\omega$ (the transition that we expect to be the most subject to environmental stochasticity in the presence of Bd) into the model described above and used it to predict long-term population viability.
Our model predicted that, following a single introduction of 40 adult individuals into a population and using the baseline value for $\omega$ of 0.3, the six populations with the highest adult survival probabilities had 50-year extinction probabilities of 0 to 0.32; @fig-viability B).
This indicates strong potential for long-term persistence in the presence of Bd and environmental variability in recruitment (@fig-viability B).
When we decreased $\omega$ from 0.3 to a conservative estimate of 0.15, 50 year extinction probability remained less than 0.5 for three populations and was 0.79 to 0.94 for three additional populations (@fig-viability B).
In contrast, for the six populations where $\sigma_{A_R} < 0.5$, extinction probability over 50 years was always predicted to be 1.
To test the validity of our model predictions, we demonstrated that our stochastic model could describe the general recovery trajectory of our translocated population with the longest survey history (@fig-viability-supp B; population 70550, surveyed for 16 years).

In summary, our model demonstrates that given observed yearly adult survival probabilities of translocated frogs, 50% of our translocated populations have a high probability of population growth and long-term viability in the presence of Bd.
This is likely a conservative estimate because there is evidence that naturally-recruited adults have higher survival probability than translocated adults (@fig-compare_surv_probs SI), but we considered these probabilities to be equal in all but three of our populations where we had sufficient data to distinguish these different probabilities.
In addition, 5 of the 7 populations translocated after 2013 were predicted to have long-term viability, compared to only 1 of 5 populations translocated prior to 2013 (@tbl-survival-earlylate).
Thus, there is some trend suggesting that our ability to select high quality habitats that promote long-term viability of translocated populations improved over time.

@@

Disease-induced population declines are impacting global biodiversity [@daszak2000], and strategies to recover affected species are lacking.
Recovery in this context is likely to be particularly challenging because typically recovery will need to occur in the presence of ongoing disease.
One potential strategy, and the one we employ in the current study, utilizes populations of resistant/tolerant individuals as a source of animals with which to reestablish extirpated populations in the presence of disease.
Although previously proposed [@mendelson2019], this strategy has rarely been successfully implemented (see [@joseph2018] for a small-scale example).
In particular, although numerous studies have documented signatures of selection in hosts following disease-induced declines [e.g., @epstein2016; @gignoux-wolfsohn2021], we are unaware of any that used individuals bearing the resulting adaptive alleles to reestablish extirpated populations.
The results of our study provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd) and that some of the genomic regions under selection are associated with genes related to disease resistance/tolerance, and (ii) reintroduction of individuals from these resistant/tolerant populations allows the successful reestablishment of viable frog populations despite ongoing Bd infection.
Although these results are generally consistent with expectations under evolutionary rescue [@carlson2014], obtaining strong evidence in wild animal populations for each of the associated causal links (i.e., selection for adaptive alleles $\rightarrow$ increase in host fitness in the presence of the pathogen $\rightarrow$ positive population growth and recovery) is difficult and perhaps impossible.
Nonetheless, in the following narrative, we use evolutionary rescue as an informative, albeit tentative framework.

Previous field studies in MYL frogs show that frog-Bd dynamics and frog survival in the presence of Bd are fundamentally different between naive and recovering populations.
Following the arrival and establishment of Bd in previously-naive populations, adult frogs develop high Bd loads that lead to mass die-offs [@vredenburg2010].
In contrast, in recovering populations adult frogs typically have low-to-moderate and relatively constant Bd loads and mass die-offs are not observed [@briggs2010; @knapp2011, see also @fig-bdload-beforeafter - before translocation].
The differences in Bd load of frogs from naive and recovering populations are also observed in controlled laboratory studies [@knapp2016, Figure 4].
The observed reduced susceptibility of frogs from recovering populations could in theory be due to several factors including natural selection for more resistant/tolerant genotypes and/or acquired immunity, but until now evidence to evaluate the role of evolution was lacking.

Results from our genomic analyses strongly suggest that natural selection for adaptive alleles and resulting increased resistance/tolerance is at least partially responsible for the reduced susceptibility of frogs in recovering populations.
We identified substantial differences in allele frequencies and larger genomic regions between adjacent Bd-naive and recovering MYL frog populations, consistent with strong selection following Bd exposure.
Importantly, some regions under selection are associated with cellular and immunological mechanisms known to contribute to disease resistance, including in amphibians [@zamudio2020].
For example, the major histocompatibility complex (MHC) plays an important role in immunity.
In our study, we identified a region that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
These results corroborate numerous previous studies linking MHC genes to amphibian resistance/tolerance against Bd [e.g., @savage2011; @bataille2015].
Similarly, the region with the strongest indication of directional selection contains the gene, GVINP1.
Several previous studies have found this gene to be differentially expressed during Bd infection [e.g., @grogan2018; @ellison2014].
This gene was also strongly linked to disease in salmon, explaining a remarkable 20% of the resistance phenotype [@robledo2020; @robledo2018].
Finally, the RIN3 gene showed a consistent pattern of allele frequency differences across all nine of the frog populations sampled in this study, indicating consistent selection in populations distributed across a wide geographic area.
In the SI, we provide additional details on these and other genes identified in our study as under selection <!--# Retain or delete? -->.
Collectively, these results suggest that the documented natural recovery of MYL frogs following Bd-caused declines [@knapp2016] is consistent with evolutionary rescue.
